---
title: "Supplemental"
#subtitle: "Introductory practical experience in conducting omics-based grouping"
authors: "Michabo Health Science Ltd"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
link-citations: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r root_dir_url, echo=FALSE}
root <- 'https://raw.githubusercontent.com/michabohealthscience/training-fsa/main'
```

# Data file summary

* **[data/HILIC_POS_male/0_sample_metadata.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/0_sample_metadata.csv)** - sample metadata
* **[data/HILIC_POS_male/0_feature_metadata.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/0_feature_metadata.csv)** - feature metadata
* **[data/HILIC_POS_male/0_sample_metadata_filtered.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/0_sample_metadata_filtered.csv)** - sample metadata (with blank removed)
* **[data/HILIC_POS_male/1_unfiltered.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/1_unfiltered.csv)** - unfiltered intensity matrix following step 1
* **[data/HILIC_POS_male/2_filtered.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/2_filtered.csv)** - filtered (both samples and features) intensity matrix following steps 1-6
* **[data/HILIC_POS_male/3_pqn.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/3_pqn.csv)** - PQN normalised intensity matrix following steps 1-7
* **[data/HILIC_POS_male/4_mv_imputed.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/4_mv_imputed.csv)** - missing value imputed intensity matrix following steps 1-8
* **[data/HILIC_POS_male/5_glog.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/HILIC_POS_male/5_glog.csv)** - glog intensity matrix following steps 1-9
* **[data/ctdf_f2000.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/ctdf_f2000.csv)** - example control vs treated dataframe for feature number 2000
* **[data/ttest_ts2.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/ttest_ts2.cs)** - t-test results for test substance 2 (both dose treatments)
* **[data/combined_male_tstats.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/combined_male_tstats.csv)** - all t-stats for all test substances and treatements for males
* **[data/unblinded_ordered_names.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/unblinded_ordered_names.csv)** - unblinded names of the test substances
* **[data/cefic_matching_compounds.csv](https://raw.githubusercontent.com/michabohealthscience/training-fsa/main/data/cefic_matching_compounds.csv)** - details of the test substances


# Mean comparison plot
Alternative plot for comparing treatment vs control
```{r compare_means, results = "hold"}
library(dplyr)
library(ggplot2)
ctdf <- read.csv('data/ctdf_f2000.csv')
ctdf %>%
    ggplot(aes(as.factor(control_treated),
               intensity)) +
    geom_jitter(aes(colour = as.factor(control_treated))) +
    stat_summary(fun.y = mean,
                 geom = "point") +
    stat_summary(fun.data = mean_se,
                 geom = "errorbar")
```


# t-test loop

Below is a function to run a t-test for all the features for a specific test substance.


```{r t-test_data_one_assay, results = "hold"}
library(reshape2)
pqn <- read.csv(file.path(root, 'data/HILIC_POS_male/3_pqn.csv'))
sample_metadata <- read.csv(file.path(root, 'data/HILIC_POS_male/0_sample_metadata_filtered.csv'))

ttest_per_test_substance <- function(pqn, sample_metadata, test_substance=2){
    pqn_t <- data.frame(t(pqn[,-1]))
    colnames(pqn_t) <- pqn$feature_name
    pqn_t$sample_name <- rownames(pqn_t)

    pqn_long <- melt(pqn_t, id.vars=c('sample_name'))
    pqn_long <- merge(pqn_long, sample_metadata[,c('sample_name', 'test_substance', 'dose_group')], on='sample_name')

    pqn_long <- pqn_long[pqn_long$test_substance != "QC",]
    pqn_long$test_substance <- as.numeric(pqn_long$test_substance)
    pqn_long$dose <- as.numeric(pqn_long$dose)
    pqn_long$value <- as.numeric(pqn_long$value)

    # Just check test substance 1
    pqn_long <- pqn_long[pqn_long$test_substance == test_substance | pqn_long$dose_group == 0, ]
    ttest_l = list()
    feature_names <- pqn$feature_name
    doses <- c(1,2)

    n = 0
    for (dose in doses){

      for (feature in feature_names){

        n = n + 1

        if (n %% 1000 == 0 || n == 1){
          print(c(n, 'of', length(feature_names)*length(doses)))
        }

        treated <- pqn_long[pqn_long$dose_group == dose &
                            pqn_long$variable == feature,]

        control <- pqn_long[pqn_long$dose_group == 0 &
                            pqn_long$variable == feature,]

        # Specify that we need 3 or more samples in both the treated and control condition
        if (sum(!is.na(treated$value)) > 2 && sum(!is.na(control$value)) > 2){
              ttest_out <- t.test(treated$value, control$value)

              if (mean(control$value, na.rm=T) > 0){
                fc <- mean(treated$value, na.rm=T) / mean(control$value, na.rm=T)
              }else{
                fc <- NA
              }


              ttest_l[[n]] <- c(feature, dose, ttest_out$p.value, ttest_out$statistic, fc)
        }else{
          ttest_l[[n]] <- c(feature, dose, NA, NA, NA)
        }

      }
    }
    ttest_df <- data.frame(Reduce(rbind, ttest_l))
    colnames(ttest_df) <- c('feature_name', 'dose', 'pvalue', 'tstat', 'fc')

    ttest_df$pvalue <- as.numeric(ttest_df$pvalue)
    ttest_df$tstat <- as.numeric(ttest_df$tstat)
    ttest_df$fc <- as.numeric(ttest_df$fc)


    ttest_df <- ttest_df[!is.na(ttest_df$pvalue),]
    rownames(ttest_df) <- NULL


    return(ttest_df)
}


```

```{r run_function, results = "hold"}
ttest_ts2 <- ttest_per_test_substance(pqn, sample_metadata, test_substance=2)
```



# t-stat - PCA

We can also represent the data using PCA

```{r PCA_unblind, fig.height=6, fig.width=8, results = "hold"}
library(ggplot2)


tstats <- read.csv(file.path(root, 'data/combined_male_tstats.csv'), row.names = 1)

tstats_scaled <- scale(tstats, center=FALSE, scale=TRUE)

unblinded_names <- read.csv(file.path(root, 'data/unblinded_ordered_names.csv'), header = TRUE)
colnames(tstats) <- unblinded_names$full_label

tstats_scaled <- scale(tstats, center=FALSE, scale=TRUE)

# transpose as features should be columns
tstats_scaled_t <- data.frame(t(tstats_scaled))

# Simply giving missing values a value of zero (but in practice this should be looked into more
# closely and missing value imputed approaches can be used
tstats_scaled_t[is.na(tstats_scaled_t)] <- 0

# remove features with no response
tstats_scaled_t_rm0 <- tstats_scaled_t[, colSums(tstats_scaled_t != 0) > 0]

# Run PCA
pca_tstats <- prcomp(tstats_scaled_t_rm0, center = TRUE, scale. = TRUE)
summ <- summary(pca_tstats)
tstats_scaled_t_rm0$full_label <- rownames(tstats_scaled_t_rm0)

pca_df <- data.frame(cbind(pca_tstats$x[,1:2], unblinded_names))

# Plot the PCA
ggplot(pca_df, aes(PC1, PC2, color=moa, label=moa)) +
   geom_text(aes(label = compound_name_dose), hjust = 0.5,  vjust = -1) +
   geom_point(aes(color=moa)) +
   stat_ellipse(geom = "polygon", aes(fill = after_scale(alpha(colour, 0.3))))+
   scale_colour_manual(values=c("orange","black","blue"))+
   xlab(paste("PC1 (", round(summ$importance[2,1]*100,1), "%)", sep='')) +
   ylab(paste("PC2 (", round(summ$importance[2,2]*100,1), "%)", sep='')) +

   theme_bw()

```


